---
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = '#',
                      fig.align = 'center', message = FALSE,
                      fig.height = 2, fig.asp = 0.5)
```

5 main steps to the analysis:

1. import data and convert to `telemetry` format, then plot by position and time

```{r, fig.asp=.9}
library('ctmm') # using the github version (0.6.1)
library('dplyr')
library('lubridate')
library('ggplot2')
theme_set(theme_bw())

tapirs <-
  read.csv('../data/1_ATLANTICFOREST_11.csv') %>%
  as.telemetry(timeformat = '%Y-%m-%d %H:%M')

joana <- tapirs[['AF_01_JOANA']] # take the first tapir

plot(joana) # default units; no information on errors, so assume DOP = 1
plot(joana, units = FALSE) # force SI units
```

```{r}
# non-stationary movement (plot is not north-oriented)
ggplot(joana, aes(x, y)) +
  geom_path(alpha = 0.2) +
  geom_point(aes(color = decimal_date(timestamp))) +
  scale_color_viridis_c('Year', breaks = c(1998:2000)) +
  labs(x = 'x (m)', y = 'y (m)')

# facet plot to see seasonal detail
ggplot(joana, aes(x, y)) +
  facet_wrap(~ year(timestamp)) +
  geom_point(aes(color = (decimal_date(timestamp) %% 1) * 12)) +
  scale_color_viridis_c('Month', breaks = c(3, 6, 9)) +
  labs(x = 'x (m)', y = 'y (m)')
```

\newpage

2. check for outliers

```{r, fig.width=4, fig.asp=.9}
diagn <- outlie(joana) # movement is not stationary
plot(diagn, units = FALSE) # very narrow CIs, nonstationary movement
```

\newpage

3. Variograms

```{r}
svf <- variogram(joana) # estimate semi-variance function
head(warnings(), n = 2) # some warnings occur occasionally
plot(svf) # plots 50% max lag by default, > 50% is "usually garbage"
plot(svf, fraction = 1e-3) # zoom in on the variance at shortest lags
```

\newpage

4. fit and select movement model

```{r, eval=FALSE, echo=FALSE}
# run this while in R to save the model but not when compiling to pdf
variogram.fit(svf, name = 'theta0', fraction = 0.0003)
fitted.mods <- ctmm.select(joana, CTMM = theta0, verbose = TRUE, cores = 0)
saveRDS(fitted.mods, 'fitted.mods.rds')
```

```{r, eval = FALSE}
variogram.fit(svf, name = 'theta0', fraction = 0.0003)
fitted.mods <- ctmm.select(joana, CTMM = theta0, verbose = TRUE, cores = 0)
```

```{r, echo = FALSE}
fitted.mods <- readRDS('fitted.mods.rds')
```

Negligible difference between OU and OUF models (both anisotropic and isotropic), but choose anisotropic since the tapir's movement is not directionally uniform.

```{r}
summary(fitted.mods)
OUa <- fitted.mods$`OU anisotropic` # Ornstein-Uhlenbeck
```

5. calculate distance and speed estimates 

```{r, fig.asp=.6}
# plot semivariance function against empirical variogram
plot(svf, CTMM = OUa, col.CTMM = '#1b9e77', fraction = 0.001)
plot(svf, CTMM = OUa, col.CTMM = '#1b9e77', fraction = 0.1)
plot(svf, CTMM = OUa, col.CTMM = '#1b9e77', fraction = 0.5)
```

\newpage

```{r, fig.height=6, fig.asp=1.25}
layout(matrix(1:6, ncol = 2))
for(i in 1:6) plot(svf, CTMM = fitted.mods[[i]], col.CTMM = 'red',
                   fraction = 2e-4, main = names(fitted.mods)[i])
for(i in 1:6) plot(svf, CTMM = fitted.mods[[i]], col.CTMM = 'red',
                   fraction = 0.01, main = names(fitted.mods)[i])
for(i in 1:6) plot(svf, CTMM = fitted.mods[[i]], col.CTMM = 'red',
                   fraction = 0.5, main = names(fitted.mods)[i])
layout(1)
summary(OUa) # parameter and estimates with CIs
speed(OUa) # OU model cannot make speed estimates (OUF can)
est <- akde(joana, CTMM = OUa)

plot(est) # 95% quantile of home range distribution with 95% CIs
plot(joana, add = TRUE) # datapoints
```

